---
title: "Assignment 5"
author: "Jamie K. Miller"
date: "November 28, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
### Load packages

library(tidyverse)
library(car)
library(stargazer)

```

```{r, include=FALSE}
### Read in data and data source

# 1. Male and Female graduate enrollment (1967 - 2015) data source

grad_enrollment_1967_2015 <- read_csv("grad_enrollment_1967_2015.csv")

# 49 obs of 4 variables

# Table 303.80. Total postbaccalaureate fall enrollment in degree-granting postsecondary institutions, by attendance status, sex of student, and control of institution: 1967 through 2015												

# NOTE: Data include unclassified graduate students. Data through 1995 are for institutions of higher education, while later data are for degree-granting institutions. Degree-granting institutions grant associate’s or higher degrees and participate in Title IV federal financial aid programs. The degree-granting classification is very similar to the earlier higher education classification, but it includes more 2-year colleges and excludes a few higher education institutions that did not grant degrees. Some data have been revised from previously published figures.													
													
# SOURCE: U.S. Department of Education, National Center for Education Statistics, Higher Education General Information Survey (HEGIS), "Fall Enrollment in Colleges and Universities" surveys, 1967 through 1985; Integrated Postsecondary Education Data System (IPEDS), "Fall Enrollment Survey" (IPEDS-EF:86-99); IPEDS Spring 2001 through Spring 2016, Fall Enrollment component; and Enrollment in Degree-Granting Institutions Projection Model, 2000 through 2026. (This table was prepared February 2017.)													
													
# Data: National Center for Education Statistics													
# https://nces.ed.gov/programs/digest/d16/tables/dt16_303.80.asp													

# reformated file and saved to csv. simplified columns
# saved as 'grad_enrollment_1967_2015'

#############################

# 2. Shifts in female PhD recipients by field (1985, 2000, 2015)

female_phd_by_field <- read_csv("female_phd_by_field.csv")

# 4 obs of 7 variables

# a Excludes respondents who did not report sex: 171 in 1995, 73 in 2000, 66 in 2005, 16 in 2010, and 7 in 2015.		# b Includes agricultural sciences and natural resources; biological and biomedical sciences; and health sciences.	# c Non-science and engineering fields not shown separately.										

# SOURCE: National Science Foundation, National Center for Science and Engineering Statistics, Survey of Earned Doctorates.										

# Table 14. Doctorate recipients, by broad field of study and sex: Selected years, 1985–2015

# reformated file and saved to csv. simplified columns
# saved as 'female_phd_by_field.csv'

#############################

# 3. Male and Female starting salaries

median_salary <- read_csv("median_salary.csv")

# 15 obs of 5 variables

# Table 48. Median basic annual salary for doctorate recipients with definite postgraduation plans in the United States, by field of study, type of postgraduation plans, and sex: 2015, (Dollars)

# NOTES: Median salaries in this table are the exact salary values of respondents at the 50th percentile of their frequency distribution; salary values have not been rounded. See table A-6 in the technical notes for a listing of major fields and their constituent subfields.									
# SOURCE: National Science Foundation, National Center for Science and Engineering Statistics, Survey of Earned Doctorates, 2015.									

# reformated file and saved to csv. simplified columns
# saved as 'median_salary.csv'

#############################

# 4. faculty salary data (2008 - 2009 survey) data source

faculty_salary_2008_2009 <- read_csv("faculty_salary_2008_2009.csv")

# 397 obs of 6 variables

# Faculty salary by rank, discipline, experience and sex
# Fox J. and Weisberg, S. (2011) An R Companion to Applied Regression, Second Edition Sage.
# The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S. The data were collected as part of the on-going effort of the college's administration to monitor salary differences between male and female faculty members.

# reformated file and saved to csv. simplified columns
# Discipline (A = Theoretical, B = Applied) changed to 'field'
# saved as 'faculty_salary_2008_2009.csv'

```
### 1. Male and Female graduate enrollment (1967 - 2015)
```{r, echo = FALSE, include = FALSE}

# Basic Linear regression for female and male graduate student enrollments 

enrollment_female_lm1 <- lm(grad_enrollment_1967_2015$total_females ~ grad_enrollment_1967_2015$Year)
#enrollment_female_lm1

enrollment_female_lm2 <- lm(total_females ~ Year, data = grad_enrollment_1967_2015)
enrollment_female_lm2

enrollment_male_lm1 <- lm(grad_enrollment_1967_2015$total_males ~ grad_enrollment_1967_2015$Year)
#enrollment_male_lm1

enrollment_male_lm2 <- lm(total_males ~ Year, data = grad_enrollment_1967_2015)
enrollment_male_lm2

# Two ways to do the exact same test.

```

```{r, include = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)
# Model diagnostics

par(mfrow = c(2,2))
plot(enrollment_female_lm2)

par(mfrow = c(2,2))
plot(enrollment_male_lm2)

# I guess the residuals are normally distributed.

```

```{r, echo = FALSE, , include = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)
# Model fit and significance

summary_f_enrollment_lm2 <- summary(enrollment_female_lm2)
summary_f_enrollment_lm2

summary_m_enrollment_lm2 <- summary(enrollment_male_lm2)
summary_m_enrollment_lm2

```

```{r, echo = FALSE, include = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)
# Pearson's *r* correlation output

pearsonsr_f_enrollment_year <- cor.test(grad_enrollment_1967_2015$total_females,
         grad_enrollment_1967_2015$Year)
pearsonsr_f_enrollment_year

pearsonsr_m_enrollment_year <- cor.test(grad_enrollment_1967_2015$total_males,
         grad_enrollment_1967_2015$Year)
pearsonsr_m_enrollment_year

```

```{r, echo = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)

enrollment_graph <- ggplot(grad_enrollment_1967_2015, aes(x = Year)) +
  geom_point(aes(y = total_females, color = "Female")) +
  geom_smooth(aes(x = Year, y = total_females), method = lm, se = TRUE, size = 0.5, color = "gray20") +
  geom_point(aes(y = total_males, color = "Male")) +
  geom_smooth(aes(x = Year, y = total_males), method = lm, se = TRUE, size = 0.5, color = "gray20") +
  theme_classic() +
  ylab("Enrolled Students\n") +
  xlab("\nYear") +
  ggtitle("Graduate Student Enrollment by Year (1967 - 2015)") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.title = element_blank())
  
enrollment_graph

```

**Figure 1. Graduate strudent enrollment by year from 1967 to 2015.** Fall enrollment of female and male students each year from 1967 to 2015.  *SOURCE: U.S. Department of Education, National Center for Education Statistics, Higher Education General Information Survey (HEGIS), "Fall Enrollment in Colleges and Universities" surveys, 1967 through 1985; Integrated Postsecondary Education Data System (IPEDS), "Fall Enrollment Survey" (IPEDS-EF:86-99); IPEDS Spring 2001 through Spring 2016, Fall Enrollment component. Prepared February 2017.*	

Graduate enrollment of female students is significantly predicted by year (*b* = 30126, t(`r pearsonsr_f_enrollment_year$parameter`) = `r round(pearsonsr_f_enrollment_year$statistic,2)`, *p* < 0.001) with a strong positive correlation between the two (Pearson's *r* = `r round(pearsonsr_f_enrollment_year$estimate,2)`. The model (*enrolled students* = 30126(*Year*) - 58955502) explains a significant amount of variance in female graduate student enrollment (F(1,47) = 2669, *p* < 0.001, $R^2$ = 0.98).

Similarly, the model (*enrolled students* = 9069(*Year*) - 17112153) explains a significant amount of variance in male graduate student enrollment (F(1,47) = 276, *p* < 0.001, $R^2$ = 0.85). Graduate enrollment of male students is also significantly predicted by year (*b* = 9069, t(`r pearsonsr_m_enrollment_year$parameter`) = `r round(pearsonsr_m_enrollment_year$statistic,2)`, *p* < 0.001) with a strong positive correlation between the two (Pearson's *r* = `r round(pearsonsr_m_enrollment_year$estimate,2)`.

Graduate enrollment generally increased for both male and female students from 1967 to 2015, though enrollment of female graduate students increased faster than enrollment of male students. In 1967 women comprised 29.6% of all enrolled graduate students. The sex ratio of graduate students switched from to male-biased to female biased in 1988 when the percentage of female students increased from 49.8% of all students in 1987 to 50.3% in 1988. By 2015, 58.5% of graduate students were female.  In recent years, enrollment leveled and even decreased slightly staring in 2010 following the 2008/2009 financial crisis.