---
title: "Assignment 5"
author: "Jamie K. Miller"
date: "November 28, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
### Load packages

library(tidyverse)
library(car)
library(stargazer)

```

```{r, include=FALSE}
### Read in data and data source

# 1. Male and Female graduate enrollment (1967 - 2015) data source

grad_enrollment_1967_2015 <- read_csv("grad_enrollment_1967_2015.csv")

# 49 obs of 4 variables

# Table 303.80. Total postbaccalaureate fall enrollment in degree-granting postsecondary institutions, by attendance status, sex of student, and control of institution: 1967 through 2015												

# NOTE: Data include unclassified graduate students. Data through 1995 are for institutions of higher education, while later data are for degree-granting institutions. Degree-granting institutions grant associate’s or higher degrees and participate in Title IV federal financial aid programs. The degree-granting classification is very similar to the earlier higher education classification, but it includes more 2-year colleges and excludes a few higher education institutions that did not grant degrees. Some data have been revised from previously published figures.													
													
# SOURCE: U.S. Department of Education, National Center for Education Statistics, Higher Education General Information Survey (HEGIS), "Fall Enrollment in Colleges and Universities" surveys, 1967 through 1985; Integrated Postsecondary Education Data System (IPEDS), "Fall Enrollment Survey" (IPEDS-EF:86-99); IPEDS Spring 2001 through Spring 2016, Fall Enrollment component; and Enrollment in Degree-Granting Institutions Projection Model, 2000 through 2026. (This table was prepared February 2017.)													
													
# Data: National Center for Education Statistics													
# https://nces.ed.gov/programs/digest/d16/tables/dt16_303.80.asp													

# reformated file and saved to csv. simplified columns
# saved as 'grad_enrollment_1967_2015'

#############################

# 2. Shifts in female PhD recipients by field (1985, 2000, 2015)

female_phd_by_field <- read_csv("female_phd_by_field.csv")

# 4 obs of 7 variables

# a Excludes respondents who did not report sex: 171 in 1995, 73 in 2000, 66 in 2005, 16 in 2010, and 7 in 2015.		# b Includes agricultural sciences and natural resources; biological and biomedical sciences; and health sciences.	# c Non-science and engineering fields not shown separately.										

# SOURCE: National Science Foundation, National Center for Science and Engineering Statistics, Survey of Earned Doctorates.										

# Table 14. Doctorate recipients, by broad field of study and sex: Selected years, 1985–2015

# reformated file and saved to csv. simplified columns
# saved as 'female_phd_by_field.csv'

#############################

# 3. Male and Female starting salaries

median_salary <- read_csv("median_salary.csv")

# 15 obs of 5 variables

# Table 48. Median basic annual salary for doctorate recipients with definite postgraduation plans in the United States, by field of study, type of postgraduation plans, and sex: 2015, (Dollars)

# NOTES: Median salaries in this table are the exact salary values of respondents at the 50th percentile of their frequency distribution; salary values have not been rounded. See table A-6 in the technical notes for a listing of major fields and their constituent subfields.									
# SOURCE: National Science Foundation, National Center for Science and Engineering Statistics, Survey of Earned Doctorates, 2015.									

# reformated file and saved to csv. simplified columns
# saved as 'median_salary.csv'

#############################

# 4. faculty salary data (2008 - 2009 survey) data source

faculty_salary_2008_2009 <- read_csv("faculty_salary_2008_2009.csv") %>%
  mutate(field_name = case_when(
    field == "A" ~ "Theoretical",
    field == "B" ~ "Applied"
  ))

# 397 obs of 6 variables

# Faculty salary by rank, discipline, experience and sex
# Fox J. and Weisberg, S. (2011) An R Companion to Applied Regression, Second Edition Sage.
# The 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S. The data were collected as part of the on-going effort of the college's administration to monitor salary differences between male and female faculty members.

# reformated file and saved to csv. simplified columns
# Discipline (A = Theoretical, B = Applied) changed to 'field'
# saved as 'faculty_salary_2008_2009.csv'

```
## 1. Male and Female graduate enrollment (1967 - 2015)
```{r, echo = FALSE, include = FALSE}

# Basic Linear regression for female and male graduate student enrollments 

enrollment_female_lm1 <- lm(grad_enrollment_1967_2015$total_females ~ grad_enrollment_1967_2015$Year)
#enrollment_female_lm1

enrollment_female_lm2 <- lm(total_females ~ Year, data = grad_enrollment_1967_2015)
enrollment_female_lm2

enrollment_male_lm1 <- lm(grad_enrollment_1967_2015$total_males ~ grad_enrollment_1967_2015$Year)
#enrollment_male_lm1

enrollment_male_lm2 <- lm(total_males ~ Year, data = grad_enrollment_1967_2015)
enrollment_male_lm2

# Two ways to do the exact same test.

```

```{r, include = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)
# Model diagnostics

par(mfrow = c(2,2))
plot(enrollment_female_lm2)

par(mfrow = c(2,2))
plot(enrollment_male_lm2)

# I guess the residuals are normally distributed.

```

```{r, echo = FALSE, , include = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)
# Model fit and significance

summary_f_enrollment_lm2 <- summary(enrollment_female_lm2)
summary_f_enrollment_lm2

summary_m_enrollment_lm2 <- summary(enrollment_male_lm2)
summary_m_enrollment_lm2

```

```{r, echo = FALSE, include = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)
# Pearson's *r* correlation output

pearsonsr_f_enrollment_year <- cor.test(grad_enrollment_1967_2015$total_females,
         grad_enrollment_1967_2015$Year)
pearsonsr_f_enrollment_year

pearsonsr_m_enrollment_year <- cor.test(grad_enrollment_1967_2015$total_males,
         grad_enrollment_1967_2015$Year)
pearsonsr_m_enrollment_year

```

```{r, echo = FALSE}
# 1. Male and Female graduate enrollment (1967 - 2015)

enrollment_graph <- ggplot(grad_enrollment_1967_2015, aes(x = Year)) +
  geom_point(aes(y = total_females, color = "Female")) +
  geom_smooth(aes(x = Year, y = total_females), method = lm, se = TRUE, size = 0.5, color = "gray20") +
  geom_point(aes(y = total_males, color = "Male")) +
  geom_smooth(aes(x = Year, y = total_males), method = lm, se = TRUE, size = 0.5, color = "gray20") +
  theme_classic() +
  ylab("Enrolled Students\n") +
  xlab("\nYear") +
  ggtitle("Graduate Student Enrollment by Year (1967 - 2015)") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.title = element_blank())
  
enrollment_graph

```

**Figure 1. Graduate strudent enrollment by year from 1967 to 2015.** Fall enrollment of female and male students each year from 1967 to 2015.  *Source: U.S. Department of Education, National Center for Education Statistics, Higher Education General Information Survey (HEGIS), "Fall Enrollment in Colleges and Universities" surveys, 1967 through 1985; Integrated Postsecondary Education Data System (IPEDS), "Fall Enrollment Survey" (IPEDS-EF:86-99); IPEDS Spring 2001 through Spring 2016, Fall Enrollment component. Prepared February 2017.*	

Graduate enrollment of female students is significantly predicted by year (*b* = 30126, t(`r pearsonsr_f_enrollment_year$parameter`) = `r round(pearsonsr_f_enrollment_year$statistic,2)`, *p* < 0.001) with a strong positive correlation between the two (Pearson's *r* = `r round(pearsonsr_f_enrollment_year$estimate,2)`. The model (*enrolled students* = 30126(*Year*) - 58955502) explains a significant amount of variance in female graduate student enrollment (F(1,47) = 2669, *p* < 0.001, R^2^ = 0.98).

Similarly, the model (*enrolled students* = 9069(*Year*) - 17112153) explains a significant amount of variance in male graduate student enrollment (F(1,47) = 276, *p* < 0.001, R^2^ = 0.85). Graduate enrollment of male students is also significantly predicted by year (*b* = 9069, t(`r pearsonsr_m_enrollment_year$parameter`) = `r round(pearsonsr_m_enrollment_year$statistic,2)`, *p* < 0.001) with a strong positive correlation between the two (Pearson's *r* = `r round(pearsonsr_m_enrollment_year$estimate,2)`.

Graduate enrollment generally increased for both male and female students from 1967 to 2015, though enrollment of female graduate students increased faster than enrollment of male students. In 1967 women comprised 29.6% of all enrolled graduate students. The sex ratio of graduate students switched from to male-biased to female biased in 1988 when the percentage of female students increased from 49.8% of all students in 1987 to 50.3% in 1988. By 2015, 58.5% of graduate students were female.  In recent years, enrollment leveled and even decreased slightly staring in 2010 following the 2008/2009 financial crisis.


## Part 4. Exploring academic salaries for professors in U.S. colleges.
#### First by Rank
```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex

# Calculate mean salary by rank
mean_by_rank <- faculty_salary_2008_2009 %>% 
  group_by(rank) %>% 
  summarize(
    mean = mean(salary)
  )

# Visualize salary by rank
by_rank <- ggplot(faculty_salary_2008_2009, aes(x = salary)) +
  geom_density(aes(color = rank, fill = rank), alpha = 0.3) +
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,250000)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "Salary (USD)", y = "Density")

by_rank

# So far, graphically, this appears to be the biggest driver.

```

#### By Sex
```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex

# Calculate mean salary by sex
mean_by_sex <- faculty_salary_2008_2009 %>% 
  group_by(sex) %>% 
  summarize(
    mean = mean(salary)
  )

# Visualize salary by sex
by_sex <- ggplot(faculty_salary_2008_2009, aes(x = salary)) +
  geom_density(aes(color = sex, fill = sex), alpha = 0.3) +
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,250000)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "Salary (USD)", y = "Density")

by_sex

```

#### By Field
```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# Discipline (A = Theoretical, B = Applied) changed to 'field'

# Calculate mean salary by Experience - Years Since PhD
mean_by_field <- faculty_salary_2008_2009 %>% 
  group_by(field_name) %>% 
  summarize(
    mean = mean(salary)
  )

# Visualize salary by Experience - Years Since PhD
by_field <- ggplot(faculty_salary_2008_2009, aes(x = salary)) +
  geom_density(aes(color = field_name, fill = field_name), alpha = 0.3) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,250000)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = "Salary (USD)", y = "Density")

by_field

```

#### By Years Experience
```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex

# Since yrs after phd and yrs service are continuous, and possibly collinear, will show them graphically.

by_yrs_phd <- ggplot(faculty_salary_2008_2009, aes(x = salary, y = years_since_phd)) +
  geom_point(aes(color = rank, pch = sex), alpha = 0.5) +
  facet_wrap(~field_name)

by_yrs_phd

# Does years service look the same?

by_yrs_service <- ggplot(faculty_salary_2008_2009, aes(x = salary, y = years_service)) +
  geom_point(aes(color = rank, pch = sex), alpha = 0.5) +
  facet_wrap(~field_name)

by_yrs_service

# Similar trend, but not the same. Might be collinear. Should test.

by_yrs_interaction <- ggplot(faculty_salary_2008_2009, aes(x = salary, y = years_service*years_since_phd)) +
  geom_point(aes(color = rank, pch = sex), alpha = 0.5) +
  facet_wrap(~field_name)

by_yrs_interaction


```

```{r, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM with all factors

faculty_salary_lm1 <- lm(salary ~ rank + field_name + sex + years_since_phd + years_service, data = faculty_salary_2008_2009)

summary(faculty_salary_lm1)

```

```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM with all factors 
# change rank reference level from AssocProf to AsstProf since it is the lowest. change from character to factor

faculty_salary_2008_2009$rank <- factor(faculty_salary_2008_2009$rank)

faculty_salary_2008_2009$rank <- fct_relevel(faculty_salary_2008_2009$rank, "AsstProf")

faculty_salary_lm2 <- lm(salary ~ rank + field_name + sex + years_since_phd + years_service, data = faculty_salary_2008_2009)

summary(faculty_salary_lm2)
```

```{r, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM - remove years_service

faculty_salary_lm3 <- lm(salary ~ rank + field_name + sex + years_since_phd, data = faculty_salary_2008_2009)

summary(faculty_salary_lm3)
```

```{r, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM - remove years_since phd

faculty_salary_lm4 <- lm(salary ~ rank + field_name + sex + years_service, data = faculty_salary_2008_2009)

summary(faculty_salary_lm4)
```

```{r, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM - for kicks, remove years experience

faculty_salary_lm5 <- lm(salary ~ rank + field_name + sex, data = faculty_salary_2008_2009)

summary(faculty_salary_lm5)
```

```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM with all factors 
# Add interaction term between yrs_service and yrs_since_phd

faculty_salary_lm6 <- lm(salary ~ rank + field_name + sex + years_since_phd + years_service + years_since_phd*years_service, data = faculty_salary_2008_2009)

summary(faculty_salary_lm6)
```

```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM with all factors  - remove sigle factors yrs_service and yrs_since_phd
# Add interaction term between yrs_service and yrs_since_phd

faculty_salary_lm7 <- lm(salary ~ rank + field_name + sex + years_since_phd*years_service, data = faculty_salary_2008_2009)

summary(faculty_salary_lm7) # This is the same exact model as lm6
```

```{r}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# LM with all factors  - remove experience

faculty_salary_lm8 <- lm(salary ~ rank + field_name + sex, data = faculty_salary_2008_2009)

summary(faculty_salary_lm8)
```


```{r, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# Model Diagnostics all factors - Model 2

plot(faculty_salary_lm2)

# Some concern with homoscedasticity of upper values

plot(faculty_salary_lm5)

# This didn't help explain anything. Only categorical variables are in this model, but still more variance at higher values than lower values.

plot(faculty_salary_lm6)

# Heteroscedasticity is a little better.

plot(faculty_salary_lm8)

```

```{r, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# AIC values for each model

sat_aic <- AIC(faculty_salary_lm2) # 9093.8
phd_aic <- AIC(faculty_salary_lm3) # 9097.2
service_aic <- AIC(faculty_salary_lm4) # 9096.8
remove_exp_aic <- AIC(faculty_salary_lm5) # 9095.5
interaction_aic <- AIC(faculty_salary_lm6) # 9088.6   Lowest AIC value
interaction_only_aic <- AIC(faculty_salary_lm7) # 9088.6
simple_aic <- AIC(faculty_salary_lm7) # 9088.6

```

```{r}
# Test collinearity

vif(faculty_salary_lm2)
vif(faculty_salary_lm6)

```


```{r stargazer1, results = 'asis', echo = FALSE, include = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# Prepare regression table

lm_table_1 <- stargazer(faculty_salary_lm2, type = "html")
lm_table_1

# For kicks, a table with all the models.

lm_table_2 <- stargazer(faculty_salary_lm2, faculty_salary_lm3, faculty_salary_lm4, faculty_salary_lm5, type = "html")
lm_table_2

```


**Table 2. Faculty salary regression results.** Results of three multiple linear regression models assessing factors impacting faculty 9 month salary (USD) from 2008 to 2009 (n = 397). Factors include faculty rank (Full, Associate, or Assistant Professor), discipline (Applied or Theoretical), sex, years since PhD was earned, and years of service. The first model includes all factors (AIC = 9093.8, Adjusted R^2^ = 0.446). The second model includes all factors plus an interaction term between years since PhD was earned and years of service (AIC = 9088.6, Adjusted R^2^ = 0.455). The third model includes only faculty rank, disciple, and sex (AIC = 9088.6, Adjusted R^2^ = 0.441). *Source: Fox J. and Weisberg, S. (2011). An R Companion to Applied Regression, Second Edition Sage.*
```{r stargazer2, results = 'asis', echo = FALSE}
# Part 4. Exploring academic salaries for professors in U.S. colleges.
# Faculty salary by rank, discipline, experience and sex
# Prepare regression table

# Possibly the final table

lm_table_3 <- stargazer(faculty_salary_lm2, faculty_salary_lm6, faculty_salary_lm8, type = "html")
lm_table_3

```


We developed three separate models with multiple linear regression to explain which factors may predict faculty 9 month salary in 2008 and 2009. Factors include faculty rank (Full, Associate, or Assistant Professor), discipline (Applied or Theoretical), sex, years since PhD was earned, and years of service (**Table 2**). 

The first model includes all factors predicting faculty salary (*salary* = 80,372.860 + 12,907.590(*rankAssocProf*) + 45,066.000(*rankProf*) - 14,417.630(*field_nameTheoretical*) + 4,783.493(*sexMale*) + 535.058(*years_since_phd*) - 489.516(*years_service*)) and explains a significant amount of variance (F(6,390) = 54.2, *p* < 0.001, Adjusted R^2^ = 0.446, AIC = 9093.8). 

Salary increased significantly with ascending rank, with Associate Professors (t(390) = 3.11, *p* = 0.002) earning about $13,000 more than Assistant Professors, and full Professors (t(390) = 10.6, *p* < 0.001) earning about 45,000 more than Assistant Professors. This is consistent with findings by the AAUP (2018). Faculty in Theoretical disciplines earned significantly less than those in Applied fields (*b* = -14418, t(390) = -6.15, *p* < 0.001). Though on average male faculty members were found to earn more than female faculty members, this factor was not a significant predictor of salary (*b* = 4784, t(390) = 1.24, *p* = 0.22), 

The two experience-related factors, years since PhD earned (*b* = 535, t(390) = 2.22, *p* = 0.027) and years service (*b* = -490, t(390) = -2.31, *p* = 0.021), were also significant predictors of salary in this model. However, the ways is which years service impacted salary was counter-intuitive. One would expect that as years since the PhD was earned and the years of service increase, presumably at the same rate, so would salary. A VIF test for collinearity suggested that these two factors may be collinear. To adjust for this we ran a second model that included all factors plus an interaction term between years since PhD was earned and years of service (*salary* = 75,594.920 + 6,140.317(*rankAssocProf*) + 34,357.940(*rankProf*) - 14,316.700(*field_nameTheoretical*) + 5,448.304(*sexMale*) + 1,032.650(*years_since_phd*) + 348.108(*years_service*) - 24.691(*years_since_phd*:*years_service*)). It shows similar trends to the first model and also explains a significant amount of variance in faculty salary (F(7,389) = 48.2, *p* < 0.001, Adjusted R^2^ = 0.455, AIC = 9088.6).

Conceptually this makes sense that both years since PhD and years service would predict an increase in salary, while the interaction term accounts for the differences in these factors. For example, two faculty members that earned their PhDs at the same time but were hired at different times may earn different salaries. In this model the years since PhD (*b* = 1032, t(389) = 3.41, *p* < 0.001) and the interaction term (*b* = -24.7, t(389) = -2.679, *p* = 0.008) were significant, and years service was not (*b* = 348, t(389) = 0.92, *p* = 0.36).

Since years of experience may be a confounding factor we ran a third model that included only faculty rank, disciple, and sex (*salary* = 81,932.220 + 13,723.420(*rankAssocProf*) + 47,403.320(*rankProf*) - 13,708.690(*field_nameTheoretical*) + 4,491.801(*sexMale*)). This model explained a significant amount of variance in faculty salary (F(4,392) = 79.2, *p* < 0.001, Adjusted R^2^ = 0.441, AIC = 9088.6), as well. In all cases, the sex of the faculty member was not a significant predictor of salary even though the male faculty members consistently earned around $5000 more than their female colleagues.







